---
title: "Final_Paper"
author: "Alex, Ethan, Zelda"
source: editor
format: 
  pdf:
    toc: true
execute: 
  echo: false
  message: false
  warning: false
---
 
```{r}
#| label: packages
#| message: false

# general use
library(tidyverse)
library(here)
library(janitor)
library(snakecase)
library(dplyr)

#Wrap Axis Label
library(scales)

#reading .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# visualizing
library(patchwork)
library(flextable)
library(ggridges)
library(NatParksPalettes)
library(paletteer)
library(ggrepel)
library(cowplot)
```

```{r}
#| label: data
birds <- read_csv(here("data", "birds.csv"))

venoco_water <- read_csv(here("data", "venoco-water.csv"))

```


```{r}
#| label: plot-aesthetics

colors <- c(
  "#0077BB",
  "#33BBEE",
  "#009988",
  "#EE7733",
  "#CC3311",
  "#EE3377",
  "#AA4499",
  "#AA4",
  "#374890",
  "#33B458",
  "#009",
  "#EE7",
  "#CC5869"
  
)

theme_set(
  theme_bw() +
    theme(text = element_text(size = 14))
)
```


```{r}
#| label: bird-wrangling


#created a new object from birds
bird_wy2025 <- birds %>% 
  # exclude repeat observations
  mutate(observation_date = as.Date(observation_date, format = "%m/%d/%y")) %>% 
  filter(observation_date > as.Date("2024-11-13") & observation_date < as.Date("2025-10-22"),
         repeat_observation == "No",
         e_bird_group %in% c("Waterfowl",
                             "New World Sparrows",
                             "Martins and Swallows",
                             "Shorebirds")) %>% 
  # filling in taxanomic information about order and families for eBird groups
  mutate(lowest_group = case_when(
    e_bird_group == "New World Sparrows" ~ "Passerellidae",
    e_bird_group == "Martins and Swallows" ~ "Hirundinidae",
    e_bird_group == "Shorebirds" ~ "Charadriiformes",
    e_bird_group == "Waterfowl" ~ "Anatidae"
  ))



# creating new object from most_abundant_groups
abundant_species_counts <- bird_wy2025 %>% 
  # grouping by speices
  group_by(season, e_bird_group, species) %>% 
  # calculating total counts for each species
  summarize(total_count = sum(count)) %>% 
  # arranging in descending order
  arrange(-total_count)
```


```{r}
#| label: water-wrangling

# Creating a new object from venoco_water
venoco_clean <- venoco_water |> 
  # filter to only include 2025 water year
  filter(wy == "2025") |> 
  # creating a new column to convert water elevation in ft to meters
  mutate(comp_level_m = (comp_level_ft + 2.96) * 0.3048) |> 
  # selecting columns of interest
  select(date, comp_level_m, temperature)


# create a new object from Venoco_clean
venoco_daily <- venoco_clean |> 
# group by date
  group_by(date) |> 
# calculate mean water level for each day
  summarize(mean_level_m = mean(comp_level_m, na.rm = TRUE)) |> 
# ungroup the data frame
  ungroup()

```


```{r}
#| label: wrangling-all-waterfowl

# creating new object from birds
 waterfowl_species<- bird_wy2025 |> 
  # filter to only include shorebords and observations in 2025 water year
  filter(e_bird_group == "Waterfowl") |> 
  # calculate total counts per species for a survey
  group_by(observation_date, species) |> 
  summarize(total = sum(count)) |> 
  ungroup() |> 
  # fill in any missing observation date and species combinations with 0
  complete(observation_date, species,
           fill = list(total = 0)) |> 
  # add a column to direct label lines
  group_by(species) |> 
  mutate(species_label = case_when(
    observation_date == min(observation_date) ~ species
  )) |> 
  ungroup() |> 
  mutate(species = fct_reorder(species, total, .fun = median, .desc = TRUE))

waterfowl_species_elevation <- waterfowl_species |>  
  left_join(venoco_daily,
  # naming shared columns between both data frames
  by = c("observation_date" = "date"))
  
```


```{r}
#| label: wrangling-all-shorebird

# creating new object from birds
 shorebird_species<- bird_wy2025 |> 
  # filter to only include shorebords and observations in 2025 water year
  filter(e_bird_group == "Shorebirds") |> 
  # calculate total counts per species for a survey
  group_by(observation_date, species) |> 
  summarize(total = sum(count)) |> 
  ungroup() |> 
  # fill in any missing observation date and species combinations with 0
  complete(observation_date, species,
           fill = list(total = 0)) |> 
  # add a column to direct label lines
  group_by(species) |> 
  mutate(species_label = case_when(
    observation_date == min(observation_date) ~ species
  )) |> 
  ungroup() |> 
  mutate(species = fct_reorder(species, total, .fun = median, .desc = TRUE))

shorebird_species_elevation <- shorebird_species |>  
  left_join(venoco_daily,
  # naming shared columns between both data frames
  by = c("observation_date" = "date"))
```


```{r}
#| label: shorebird-wrangling

shorebird_total_count <- bird_wy2025 |> 
  #filter to only include herons, ibis and allies
  filter(lowest_group == "Charadriiformes") |> 
  #group by species
  group_by(species) |> 
  #calc total counts per species
  summarize(total_count = sum(count)) |> 
  #arrange in descending order by total counts
  arrange(-total_count)
```


```{r}
#| label: waterfowl-wrangling

waterfowl_total_count <- bird_wy2025 |> 
  #filter to only include herons, ibis and allies
  filter(lowest_group == "Anatidae") |> 
  #group by species
  group_by(species) |> 
  #calc total counts per species
  summarize(total_count = sum(count)) |> 
  #arrange in descending order by total counts
  arrange(-total_count)
```


```{r}
#| label: wrangling-shorebirds-waterfowl

shorebird_waterfowl_25 <- bird_wy2025 |> 
  #filter to only include herons, ibis and allies
  filter(e_bird_group %in% c("Waterfowl", "Shorebirds")) |> 
  #group by species
  group_by(e_bird_group, observation_date) |> 
  #calc total counts per species
  summarize(total_count = sum(count)) |> 
  ungroup() |> 
  left_join(venoco_daily,
  # naming shared columns between both data frames
  by = c("observation_date" = "date"))


shorebird_wy2025 <- shorebird_waterfowl_25 %>% 
  filter(e_bird_group == "Shorebirds")

waterfowl_wy2025 <- shorebird_waterfowl_25 %>% 
  filter(e_bird_group == "Waterfowl")

```

```{r}
#| label: Shorebirds-Pearson-Correlation

cor.test(~ total_count + mean_level_m, data = shorebird_wy2025, method = "pearson")

```

```{r}
#| label: Waterfowl-Pearson-Correlation

cor.test(~ total_count + mean_level_m, data = waterfowl_wy2025, method = "pearson")

```


# Introduction


# Methods


# Results

First and foremost, we explored water elevation and taxonomic abundance as separated variables over time. The first plot shows water elevation through time in the 2025 water year, along with major rain events that significantly spiked water elevation. A limitation for this data is that this water elevation is accurate to local Venoco bridge water elevation, not necessarily the whole wetland.
```{r}
#| label: Venoco-water-elevation
#| fig-height: 4
#| fig-width: 10
#| 
ggplot(data = venoco_daily,
       # x axis is species in descending order of total count
       # # y axis is total count
       # filling columns by species
       mapping = aes(x = date,
                     y = mean_level_m)) + 
  #first layer: column to represent counts
  geom_point() +
  # adding line
  geom_line() +
  # adjusting appearance of x-axis
  scale_x_date(
    # setting limits (no upper bound)
    limits = c(as_date("2024-10-01"), NA),
    # setting breaks so that date labels appear every month
    breaks = as_date(c("2024-10-01", "2024-11-01", "2024-12-01",
                       "2025-01-01", "2025-02-01", "2025-03-01",
                       "2025-04-01", "2025-05-01", "2025-06-01",
                       "2025-07-01", "2025-08-01", "2025-09-01")),
    # label dates with month abbreviations
    date_labels = "%b") +
   scale_y_continuous(expand = expansion(mult = c(0.01, 0.1))) +
  # custom theme
  theme_bw() +
  # taking out the legend
  theme(legend.position = "none") +
  #relabelling x and y axes, adding title
  labs(x = "Date",
       y = "Mean Water Level (meter)",
       title = "Water Elevation through 2025") +
  # add vertical line for major rain event in January
  geom_vline(xintercept = as_date("2025-01-26"),
             linetype = "dashed") +
  # text label (from cowplot package)
  draw_label("First major rain event\n(3.40 in)",
             # x- and y-axis position
             x = as_date("2025-01-20"),   
             y = 2.6,
             # right alignment
             hjust = 1,
             # text size
             size = 10) +
  # add vertical line for major rain event in March
  geom_vline(xintercept = as_date("2025-03-12"),
             linetype = "dashed") +
  # text label
  draw_label("Last major rain event\n(1.11 in)",
             # x- and y-axis position
             x = as_date("2025-03-17"),   
             y = 2.6,
             # left alignment
             hjust = 0,
             # text size
             size = 10)
```

Here you can observe that the highest elevation reached around 2.6 meters with highest elevations concentrating in February 2025. After April, the water levels exhibit a steady decline until August. We then examined how bird counts looked over time within the two taxonomic groups Waterfowl and Shorebirds.

```{r}
#| label: Shorebird-waterfowl-time-plot
#| fig-height: 4
#| fig-width: 10

ggplot(data = shorebird_waterfowl_25,
       aes(x = observation_date,
           y = total_count,
           color = e_bird_group)) +
  # setting point geometry
  geom_point() +
  # adding line geometry
  geom_line() +
   # adjusting appearance of x-axis
  scale_x_date(
    # setting limits (no upper bound)
    limits = c(as_date("2024-10-01"), NA),
    # setting breaks so that date labels appear every month
    breaks = as_date(c("2024-10-01", "2024-11-01", "2024-12-01",
                       "2025-01-01", "2025-02-01", "2025-03-01",
                       "2025-04-01", "2025-05-01", "2025-06-01",
                       "2025-07-01", "2025-08-01", "2025-09-01")),
    # label dates with month abbreviations
    date_labels = "%b") +
   scale_y_continuous(expand = expansion(mult = c(0.01, 0.1))) +
  # changing background theme
  theme_bw() +
  # custom colors
  scale_color_paletteer_d("nationalparkcolors::DeathValley") +
  # Adding axis titles, title and legend
  labs(
    x = "Time",
    y = "Bird Count",
    color = "Taxonomic Group",
    title = "Shorebirds and Waterfowl Counts in Water Year 2025") +
  # add vertical line for major rain event in January
  geom_vline(xintercept = as_date("2025-01-26"),
             linetype = "dashed") +
  # text label (from cowplot package)
  draw_label("First major rain event\n(3.40 in)",
             # x- and y-axis position
             x = as_date("2025-01-20"),   
             y = 160,
             # right alignment
             hjust = 1,
             # text size
             size = 10) +
  # add vertical line for major rain event in March
  geom_vline(xintercept = as_date("2025-03-12"),
             linetype = "dashed") +
  # text label
  draw_label("Last major rain event\n(1.11 in)",
             # x- and y-axis position
             x = as_date("2025-03-17"),   
             y = 160,
             # left alignment
             hjust = 0,
             # text size
             size = 10)
  
```

Around the February, Waterfowl showed the highest abundance, adapting well to the high water elevation while its counterpart, the Shorebirds, showed a heavy decline in abundance from high water elevation. Shorebirds continued to decline in numbers, showing a record low right before June before rising back up through to August as water elevation continued to decline. 

After reaching their record highest abundance in February, Waterfowl began to decline in March, but suddenly shot back up in abundance in June when Shorebird abundance was almost zero. The plot highlights sharp deviation between the two taxonomic groups, possibly revealing that water elevation affects them differently.

After plotting each variable separately, we continued our exploration by creating a scatter plot of the two variables. 

```{r}
#| label: Shorebird-waterfowl-plotting
#| fig-height: 4
#| fig-width: 10

ggplot(data = shorebird_waterfowl_25,
       aes(x = mean_level_m,
           y = total_count,
           color = e_bird_group)) +
  # setting point geometry
  geom_point() +
  # adding a linear regression of best fit with the Lowess technique
  geom_smooth(method = "lm", se = TRUE) +
  # changing background theme
  theme_bw() +
  # custom colors
  scale_color_paletteer_d("nationalparkcolors::DeathValley") +
  # Adding axis titles, title and legend
  labs(
    x = "Venoco Bridge Water Elevation (meters)",
    y = "Bird Count",
    color = "Taxonomic Group",
    title = "Water Elevation Effects on Shorebirds and Waterfowl in Water Year 2025")
  
```

This plot revealed correlation between water elevation and these taxonomic groups in NCOS. We applied the Pearson's correlation test to find exact values of this correlation. Waterfowl show a positive correlation between bird abundance and water elevation with a correlation value of 0.785. In contrast, Shorebirds have a negative correlation value of -0.445. Therefore, this scatter plot shows how as water elevation increases, Shorebird abundance drops, and Waterfowl abundance climbs.


# Discussion


# References